# Introduction to the WGCNA Rpackage {#WGCNA}

## WGCNA Rpackage Installation procedure.  

**Installing and re-installing new versions or R and Studio.**  
In Terminal (or Rstudio > Terminal).  
```
brew reinstall --cask r
brew reinstall --cask rstudio
```

**Installing WGCNA: and R package for weighted correlation network analysis.**  

To install the package, go to Tab: Packages > Install …[]… -Repository: (CRAN); -Packages: WGCNA; *Install.  In R code:  

```{r, eval=FALSE}
install.packages("WGCNA")
```

**Installation (Failed)**.  
Unfortunately, it could not install correctly. **Error** was:   

```{ruby, eval= FALSE} 
install.packages("WGCNA")
Warning in install.packages :
dependencies ‘Biobase’, ‘impute’, ‘preprocessCore’, ‘GO.db’, ‘AnnotationDbi’ are not available
Package which is only available in source form, and may need compilation of C/C++/Fortran: ‘WGCNA’
Do you want to attempt to install these from sources? (Yes/no/cancel) y
Warning in install.packages :
installation of package ‘WGCNA’ had non-zero exit status
```  

Dependencies are old and require special installation.  

1. From [this page]( http://pklab.med.harvard.edu/scw2014/WGCNA.html), installed several packages. **(Failed)**.   
```{r, eval=FALSE}
install.packages(c("dynamicTreeCut", "cluster", "flashClust", "Hmisc", "reshape", "foreach", "doParallel") ) 
source("http://bioconductor.org/biocLite.R") 
biocLite("impute")
install.packages("WGCNA")
```   
Install most of the packages, but not “impute”, result: could not install WGCNA.  

1. Tried to install dependencies independently from [here](https://support.bioconductor.org/p/120430/ ): **(Failed)**    
```{eval=FALSE}
You're using R-3.6.0, but trying to install Bioconductor version 3.8. Try using BiocManager::install("Biobase"). If that works, then try BiocManager::install("GO.db").
If you have success, make sure your installation is valid with BiocManager::valid().
BiocManager::install("Biobase")
BiocManager::install("GO.db")
BiocManager::valid()
No resutl
```    
1. Decided to install the complete “BiocManager” pack. ** BiocManager: (OK); (WGCNA:(Failed)**   
```{r, eval=FALSE}
> install.packages("BiocManager") 
 downloaded 315 KB
> install.packages("WGCNA")
 Warning in install.packages : dependencies ‘impute’, ‘preprocessCore’ are not available
```

1. Search ‘preprocessCore’, in this [page]( https://bioconductor.org/packages/release/bioc/html/preprocessCore.html). **PreprocessCore:(OK); (WGCNA:(Failed)**   
To install this package, start R (version "4.1") and enter:  

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
 install.packages("BiocManager")
BiocManager::install("preprocessCore")
```

1. For ‘impute’, found install in this [page]( https://bioconductor.org/packages/release/bioc/html/impute.html ). **Impute:(OK); WGCNA:( OK)**  
To install this package, start R (version "4.1") enter:   

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
 install.packages("BiocManager")
BiocManager::install("impute")
 Then install WGCNA
install.packages("WGCNA")
```

## Final installation workflow:   

```{r, eval=FALSE}
# Set working directory
getwd()
setwd("/Users/marcelorosales/Box Sync/Documents/R/Rmarkdown")  
getwd()

# Load list of packages and install.
load("Rpackages")
for (p in setdiff(packages, installed.packages()[,"Package"]))
  install.packages(p)

Check <- installed.packages()

BiocManager::install("Biobase")
BiocManager::install("GO.db")
BiocManager::valid()
> install.packages("BiocManager")
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("preprocessCore")
BiocManager::install("WGCNA")


install.packages(c("dynamicTreeCut", "cluster", "flashClust", "Hmisc", "reshape", "foreach", "doParallel") ) 
source("http://bioconductor.org/biocLite.R") 
biocLite("impute")

if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("impute")

install.packages("WGCNA")
```

**Checking the packages for WGCNA installed previously for OS Catalina**.   

> Warning in install.packages :  package ‘anRichment’ is not available for this version of R  

A version of this package for your version of R might be available elsewhere, see the ideas at
https://cran.r-project.org/doc/manuals/r-patched/R-admin.htmlInstalling-packages

Warning in install.packages :  
  package ‘anRichment’ is not available for this version of R  
  package ‘anRichmentMethods’ is not available for this version of R  
  package ‘BiocFileCache’ is not available for this version of R  
  package ‘BiocParallel’ is not available for this version of R  
  package ‘biomaRt’ is not available for this version of R  
  package ‘DelayedArray’ is not available for this version of R  
  package ‘GenomicAlignments’ is not available for this version of R  
  package ‘GenomicFeatures’ is not available for this version of R  
  package ‘GenomicRanges’ is not available for this version of R  
  package ‘inserttable’ is not available for this version of R  
  package ‘lorem’ is not available for this version of R  
  package ‘org.Hs.eg.db’ is not available for this version of R  
  package ‘org.Mm.eg.db’ is not available for this version of R  
  package ‘Rhtslib’ is not available for this version of R  
  package ‘Rsamtools’ is not available for this version of R  
  package ‘rtracklayer’ is not available for this version of R  
  package ‘SummarizedExperiment’ is not available for this version of R  
  package ‘TxDb.Hsapiens.UCSC.hg19.knownGene’ is not available for this version of R  
  package ‘TxDb.Mmusculus.UCSC.mm10.knownGene’ is not available for this version of R  
A version of this package for your version of R might be available elsewhere,  
see the ideas at  
https://cran.r-project.org/doc/manuals/r-patched/R-admin.htmlInstalling-packages


## WGCNA tutorial.   

This code has been adapted from the tutorials available at WGCNA website **(this page does no longer exist)**.  

WGCNA: an R package for weighted correlation network analysis [site]( https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/WGCNAIntro ).   

 Tutorials for the WGCNA package. [R tutorial]( https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/Tutorials/index.html )  
Steps Required for this process are:  
1. **Data input and cleaning**: [PDF document]( https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/Tutorials/FemaleLiver-01-dataInput.pdf ), [R script]( https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/Tutorials/FemaleLiver-01-dataInput.R ).  
1. Network construction and module detection  
a. Automatic, one-step network construction and module detection: PDF document, R script  
b. Step-by-step network construction and module detection: PDF document, R script  
c. Dealing with large datasets: block-wise network construction and module detection: PDF document, R script  
1. Relating modules to external clinical traits and identifying important genes: PDF document, R script  
1. Interfacing network analysis with other data such as functional annotation and gene ontology PDF document, R script  
1. Network visualization using WGCNA functions: PDF document, R script  
1. Export of networks to external software: PDF document, R script  

```{r, eval=FALSE}

```

### 3.4 Gene expression analysis
The level of gene expression is measured by read density, the higher the read density, the higher the level of gene expression.  
Gene expression calculation was performed with the formula below, which calculates **FPKM (Fragments per kilo bases per million reads)** based on read counts from HT-seq (V 0.6.1) (Mortazavi, 2008).  

The formula is:  


Figure 3.4.1

The ratio of (total exon fragments / mapped reads [millions]) is the read count mapped to the gene normalized to total read counts. The value is then normalized to gene length (exon length [KB]), so that the expression of genes with different sequencing depths and length are comparable.

The numbers of genes with different expression levels are summarized in Table 3.7.1. In general, FPKM threshold for gene expression is set between 0.1-1, although there is no absolute standard and various thresholds have been used in the literature.















